---
title: "Time Series model fitting of Boston Weather Data"
author: "Vivek Vikram Magadi"
date: "April 13, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r, echo=FALSE,message=FALSE,warning=FALSE}
library(lubridate)
library(rugarch)
library(vars)
library(readr)
library(forecast)
library(astsa)
#library(MTS)
library(tseries)
library(gridExtra)
```

## Reading Data
The dataset that we have considered to work on is the Boston Weather dataset from July 2012 - August 2015. The dataset contains 5 different variables:
1. Temperature
2. Dewpoint
3. Humidity
4. Wind
5. Precipitation

As we shall see in the later part of the project, the precipitation variable shall not be considered as it is white noise.

Our data contains 1157 observations and 5 variables. Each observation is recorded on a daily basis. 

Hypothesis : Our hypothesis is to forecast values of each variable.
We have built two sets of models for comparison purposes. 
In the first set of models we model each variable as a univariate time series using ARIMA/SARIMA.
In the second set of models we model a Vector Auto-Regressive model on all 4 variables.
Then we compare to see if the Vector Auto-Regressive model gave us a better fit or did the Arima/Sarima models give us a better fit.

```{r , echo=FALSE,message=FALSE,warning=FALSE}
weatherdata <- read_csv("C:/Vivek's Stuff/Classes/Time Series/Project/weatherdata.xls")
str(weatherdata)

```

The timestamp was split into day, month and year variables starting from August 2015 to July 2012. we shall club them into a single variable and then reverse the order of the observations.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
weatherdata$time <- as.Date(with(weatherdata, paste(year, month, day,sep="-")), "%Y-%m-%d")
weatherdata$day <- NULL
weatherdata$month <- NULL
weatherdata$year <- NULL
#Order in reverse. 
weatherdata <- weatherdata[order(weatherdata$time),]
```


## Mean Difference
First we subtract the mean from all the variables to center them on the X axis.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
humidity <- mean(weatherdata$humidity)
precipitation <- mean(weatherdata$precipitation)
wind <- mean(weatherdata$wind)
dewpoint <- mean(weatherdata$dewpoint)
temperature <- mean(weatherdata$temperature)
```


```{r, echo=FALSE,message=FALSE,warning=FALSE}
#Subtracting mean from each of the time series
weatherdata$humidity <- weatherdata$humidity - humidity
weatherdata$precipitation <- weatherdata$precipitation - precipitation
weatherdata$wind <- weatherdata$wind - wind
weatherdata$dewpoint <- weatherdata$dewpoint - dewpoint
weatherdata$temperature <- weatherdata$temperature - temperature
```

##Plotting all the time series

```{r, echo=FALSE,message=FALSE,warning=FALSE}
tsplot(weatherdata$precipitation,x = weatherdata$time,main="Precipitation")
tsplot(weatherdata$temperature,x = weatherdata$time,main="Temperature")
tsplot(weatherdata$dewpoint,x = weatherdata$time,main="Dewpoint")
tsplot(weatherdata$humidity,x = weatherdata$time,main="Humidity")
tsplot(weatherdata$wind,x = weatherdata$time,main="Wind")
```

From the above Time Series plots we can see that precipitation almost looks like white noise; Temperature and Dewpoint have seasonal patterns ( looks like a yearly pattern).

## Decomposing Time Series plots
```{r, echo=FALSE,message=FALSE,warning=FALSE}
#Temperature
temperature_stl<- stl(ts(weatherdata$temperature,frequency=365),s.window = "periodic")
plot(temperature_stl,main="Temperature")
#Humidity
humidity_stl<- stl(ts(weatherdata$humidity,frequency=30),s.window = "periodic")
plot(humidity_stl,main="Humidity")
#Dewpoint
dewpoint_stl<- stl(ts(weatherdata$dewpoint,frequency=365),s.window = "periodic")
plot(dewpoint_stl,main="Dewpoint")
#Wind
wind_stl<- stl(ts(weatherdata$wind,frequency=30),s.window = "periodic")
plot(wind_stl,main="Wind")
#Precipitation
precipitation_stl <- stl(ts(weatherdata$precipitation,frequency=30),s.window = "periodic")
plot(precipitation_stl,main="Precipitation")
```

The above plots show a decomposed version of each time series by using loess. Temperature and Dewpoint are considered on a yearly basis whereas humidity, wind and precipitation are considered on a monthly basis.


## Interpretting ACF and PACF 

## Humidity
```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(5,5,5,5))
acf(weatherdata$humidity,plot = T,main="ACF for Humidity")
pacf(weatherdata$humidity,plot=T,main=" PACF for humidity")
```

We can see that the lags in AutoCorrelation and Partial Autocorrelation plots die off after a certain number of significant lags so we can assume stationarity. To quantify stationarity we shall conduct the dickey-fuller test.

By looking at the significant lags of the ACF we determine the MA lag term and that of the PACF determines the ACf term.
ACF : 3 significant lags. Also there looks like there exists a seasonal pattern in the data as denoted by lag 12 and it's surroundings.
PACF : 2 significant lags. The seasonal lags persist in the PACF plot as well.


## Dewpoint
```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))
acf(weatherdata$dewpoint,plot=T,main="ACF for Dewpoint")
pacf(weatherdata$dewpoint,plot=T,main="PACF for Dewpoint")
```

The PACF plot indicates that a significant correlation occurs at lag 2.
The ACF plot indicates that an auto-regressive model might be a right fit to the model as indicated by the slow decay in the correlations.

## Precipitation

```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(3,3,3,3))
acf(weatherdata$precipitation,plot=T,main="ACF for Precipitation")
pacf(weatherdata$precipitation,plot=T,main ="PACF for Precipitation")
```

The ACF plot does not present any significant lags. But it is noticed that lags of same correlation occur at regular intervals of 4.
The PACF plot shows that the significant correlations occur only at lags 3 and 22.
This series mainly indicates white noise and shall not be considered for our analysis.

## Wind
```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))
acf(weatherdata$wind,plot=T,main="ACf for Wind")
pacf(weatherdata$wind,main="PACF for wind")
```

The ACf tells us that there is a significant correlation at lag 1. Also a seasonal lag of 5 might be present.
The PACF does not present us with much information apart from the fact that some seasonal lag might exist.

## Temperature
```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))
acf(weatherdata$temperature,plot=T,main = "ACF for temperature")
pacf(weatherdata$temperature,plot=T,main ="PACF for temperature")
```

The ACF plot shows slow decay indicating an auto-regressive model might be appropriate over here. Looking at the PACF we can see that a significant correlation occurs at lag 2 followed by some seasonal variation. 

##Spectral Analysis



```{r, echo=FALSE,message=FALSE,warning=FALSE}
spec.pgram(weatherdata$temperature,spans=c(5,5),main = "Temperature")
spec.pgram(weatherdata$wind,log="no",spans=c(5,5),main = "Wind")
spec.pgram(weatherdata$dewpoint,spans=c(5,5),main = "Dewpoint")
spec.pgram(weatherdata$humidity,log="no",spans=c(10,10),main = "Humidity")
```

Except wind all the other variables have the trend that as the frequency gets higher, the amplitude gets smaller.
Humidity shows that it the frequency with the highest amplitude indicates that it changes every ~90 days(1 season).
Wind also looks like the frequency with the highest amplitude occurs at a cycle of about 80 - 90 days. 
Temperature and Dewpoint spectral plots are similar with dewpoint having a higher amplitude at corresponding frequencies.

## Fitting Data

We noticed that there was seasonality in temperature and dewpoint We shall get rid of them by harmonic regression. A frequency of 3/N was taken.
Temperature.


```{r, echo=FALSE,message=FALSE,warning=FALSE}
#Temperature
a <- length(weatherdata$temperature)
time <- 1:a
s1 <- cos(2*3.14*time*(3/a))
s2 <- sin(2*3.14*time*(3/a))
temp_res <- lm(weatherdata$temperature~time +s1)
tsplot(temp_res$fitted.values,main = "Fitted Values of Harmonic Regresison")

#Check residuals 
tsplot(temp_res$residuals,main="Residuals of Harmonic Regresison")

```

Dewpoint

```{r, echo=FALSE,message=FALSE,warning=FALSE}
#Dewpoint

s12 <- cos(2*3.14*time*(3/a))
s22 <- sin(2*3.14*time*(3/a))
temp_dew <- lm(weatherdata$dewpoint~time +s12+s22)
tsplot(temp_dew$fitted.values,main="Fitted values of dewpoint after harmonic regression")

#Check Residuals
tsplot(temp_res$residuals,main="Residuals after harmonic Regression")

```

We shall work on the residuals of harmonic regression applied on temperature and dewpoint variables.


##Checking stationarity via Dickey - Fuller Test

Dickey- fuller test checks the null hypothesis whether unit roots of the characteristic equation is present. If the p-value of the test is below 0.05 we reject the null hypothesis in favor of the alternative saying that all the roots of the charactersitic equation are greater than 1 ie the roots lie outside the unit circle.

Humidity

```{r, echo=FALSE,message=FALSE,warning=FALSE}
adf.test(weatherdata$humidity,alternative = "stationary")
```


Dew Point

```{r, echo=FALSE,message=FALSE,warning=FALSE}
adf.test(temp_dew$residuals,alternative = "stationary")
```

Wind

```{r, echo=FALSE,message=FALSE,warning=FALSE}
adf.test(weatherdata$wind,alternative = "stationary")
```

Temperature

```{r, echo=FALSE,message=FALSE,warning=FALSE}
adf.test(temp_res$residuals,alternative = "stationary")
```

All of the p-values are significant and hence we conclude that all of our time series are stationary.

## Cross Correlation Plots



```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))
ccf(weatherdata$humidity,temp_res$residuals,main="Humidity and Temperature")
```

Humidity lags temperature. Thus humidity can be used as a predictor for temperature.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))
ccf(weatherdata$humidity,temp_dew$residuals,main="Humidity and Dewpoint")
```

Humidity slightly leads dewpoint. Dewpoint might be a useful predictor for Humidity.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))

ccf(weatherdata$humidity,weatherdata$wind,main="Humidity and Wind")
```

Humidity slightly leads wind. Wind might be a useful predictor for humidity.




```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))

ccf(temp_res$residuals,temp_dew$residuals,main="Dewpoint and Temperature")
```

There is no strict lag/lead relationship that we can infer from this plot. We can just say that both the variables cause each other just like a cycle.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))

ccf(temp_dew$residuals,weatherdata$wind,main="Dewpoint and Wind")
```

Dewpoint lags wind at lag 1,2 and 3.
Dewpoint might be a useful predictor for wind. 

```{r, echo=FALSE,message=FALSE,warning=FALSE}
par(mar=c(4,4,4,4))

ccf(weatherdata$wind,temp_res$residuals,main="Wind and Temperature")
```

Wind leads Temperature at lags 2,3 and 4.
Temperature might be a useful predictor for wind.


Now all the time series are stationary

##Model Fitting

Going to fit an ARIMA model for each of them to see how it performs.

Temperature

```{r, echo=FALSE,message=FALSE,warning=FALSE}
fit_temp <- Arima(temp_res$residuals,order=c(2,0,2),seasonal = list(order=c(1,0,0),period=12))
tsplot(weatherdata$temperature,main = "Temperature fit")
lines(temp_res$fitted.values +fit_temp$fitted,col="red")
fit_temp
par(mar=c(4,4,4,4))
acf(fit_temp$residuals,plot=T,main="Residuals of Temperature")
```

Dewpoint

```{r, echo=FALSE,message=FALSE,warning=FALSE}
fit_dew <- Arima(temp_dew$residuals,order=c(2,0,2),seasonal = list(order=c(1,0,0),period=12))
tsplot(weatherdata$dewpoint,main ="Dewpoint Fit")
lines(temp_dew$fitted.values + fit_dew$fitted,col="red")
fit_dew
par(mar=c(4,4,4,4))
acf(fit_dew$residuals,plot=T,main="Residuals of Dewpoint")
```

Humidity

```{r, echo=FALSE,message=FALSE,warning=FALSE}
fit_hum <- Arima(weatherdata$humidity,order=c(1,0,2))
tsplot(weatherdata$humidity,main="Humidity Fit")
lines(fit_hum$fitted,col="red")
fit_hum
par(mar=c(4,4,4,4))
acf(fit_hum$residuals,plot=T,main="Residuals of Humidity")
```

Wind

```{r, echo=FALSE,message=FALSE,warning=FALSE}
fit_win <- Arima(weatherdata$wind,order=c(1,0,2))
tsplot(weatherdata$wind,main = "Wind Fit")
lines(fit_win$fitted,col="red")
fit_win
par(mar=c(4,4,4,4))
acf(fit_win$residuals,plot=T,main="Residuals of Wind")
#plot(predict(fit_win,n.ahead=10))
```

Conclusion : Temperature and Dewpoint models presented a good fit. Whereas humidity model fit was underfit and the wind model fit was overshot.

##VAR Modelling

```{r, echo=FALSE,message=FALSE,warning=FALSE}
newdata <- data.frame(temperature=temp_res$residuals,dewpoint=temp_dew$residuals,wind=weatherdata$wind,humidity=weatherdata$humidity)

VARselect(weatherdata[,c(2:5)])


var_fit <- VAR(weatherdata[,c(2:5)],p = 5,type="const")
#var_pred <- predict(var_fit,newdata,n.ahead = 100)
#plot(var_pred)
```

Model Parameters

```{r,echo=F,message=F,warning=F}
summary(var_fit)
```


```{r,echo=F,warning=F,message = F}

#checking fit
#Temperature
tsplot(weatherdata$temperature,main = "Temperature")
lines(fitted(var_fit)[,1],col="red")

#Humidity
tsplot(weatherdata$humidity,main ="Humidity")
lines(fitted(var_fit)[,3],col="red")

#Wind
tsplot(weatherdata$wind,main = "Wind")
lines(fitted(var_fit)[,4],col="red")


#Dewpoint
tsplot(weatherdata$dewpoint,main = "Dewpoint")
lines(fitted(var_fit)[,2],col="red")


#Test residuals
par(mar=c(4,4,4,4))
#Temperature
acf(resid(var_fit)[,1],plot=T,main="Temperature Residuals")
Box.test(resid(var_fit)[,1],lag=5)
#Dewpoint
acf(resid(var_fit)[,2],plot=T,main = "Dewpoint Residuals")
#Humidity
acf(resid(var_fit)[,3],main = "Humidity Residuals")
#Wind
acf(resid(var_fit)[,4],main ="Wind Residuals")


#Plotting Actual residuals
#tsplot(resid(var_fit)[,1])
#tsplot(resid(var_fit)[,2])
#tsplot(resid(var_fit)[,3])
#tsplot(resid(var_fit)[,4])

```

Granger - Causality
```{r,echo=F,warning=F,message=F}
causality(var_fit,cause ="dewpoint")$Granger
```

Temperature has an effect on dewpoint,humidity and wind.

Wind has an effect on temperature, dewpoint and humidity.

Humidity does not have a significant effect on temperature,dewpoint and wind.

Dewpoint(p-value : 0.06) may or may not have an effect on temperature,humidity and wind depending on the threshold.

CCF Results:
Humidity might be used to predict wind and temperature.

Temperature can be predicted by Humidity.

Dewpoint might be a useful predictor for Humidity.

Dewpoint might be a good predictor for wind.

Dewpoint and temperature seem to have a cycle effect.

Temperature might be a useful predictor for wind.

## Extra: Forecasting using STL 

Using STL Decompostion.
Fit ARIMA model on residuals.



Temperature:
  Using STL Decomposition + ARIMA on the residuals.
Frequency is set to 365 as seasons occur on a yearly basis.

```{r, echo=FALSE,message=FALSE,warning=FALSE}
temperature_stl<- stl(ts(weatherdata$temperature,frequency=365),s.window = "periodic")
temperature_df <- data.frame(temperature_stl$time.series)

#Finding out AR and MA lags for the remainder of the series
acf(temperature_df$remainder)
pacf(temperature_df$remainder)

fit <- Arima(temperature_df$remainder,order=c(1,0,1))
AIC(fit)
BIC(fit)

fit

f <- forecast(temperature_stl, h=250,forecastfunction=function(x,h,level){
  fit <- Arima(x,order=c(1,0,1),include.mean = F)
  return(forecast(fit,h=250,level=level))})
plot(f,include=1000)
```

Humidity:
  
```{r, echo=FALSE,message=FALSE,warning=FALSE}
  
humidity_stl<- stl(ts(weatherdata$humidity,frequency=365),s.window = "periodic")
humidity_df <- data.frame(humidity_stl$time.series)


#Finding out AR and MA lags for the remainder of the series
acf(humidity_df$remainder)
pacf(humidity_df$remainder)

fit <- Arima(humidity_df$remainder,order=c(1,0,1))
AIC(fit)
BIC(fit)

fit
f <- forecast(humidity_stl, h=250,forecastfunction=function(x,h,level){
  fit <- Arima(x,order=c(1,0,1),include.mean = F)
  return(forecast(fit,h=250,level=level))})
plot(f,include=1000)
```


Dewpoint
```{r, echo=FALSE,message=FALSE,warning=FALSE}
dewpoint_stl<- stl(ts(weatherdata$dewpoint,frequency=365),s.window = "periodic")
dewpoint_df <- data.frame(dewpoint_stl$time.series)

#Finding out AR and MA lags for the remainder of the series
acf(dewpoint_df$remainder)
pacf(dewpoint_df$remainder)

fit

fit <- Arima(dewpoint_df$remainder,order=c(1,0,1))
AIC(fit)
BIC(fit)

f <- forecast(dewpoint_stl, h=250,forecastfunction=function(x,h,level){
  fit <- Arima(x,order=c(1,0,1),include.mean = F)
  return(forecast(fit,h=250,level=level))})
plot(f,include=1000)
```


Wind
```{r, echo=FALSE,message=FALSE,warning=FALSE}
wind_stl<- stl(ts(weatherdata$wind,frequency=365),s.window = "periodic")
wind_df <- data.frame(wind_stl$time.series)

#Finding out AR and MA lags for the remainder of the series
acf(wind_df$remainder)
pacf(wind_df$remainder)

fit <- Arima(wind_df$remainder,order=c(0,0,1))
AIC(fit)
BIC(fit)

fit

f <- forecast(wind_stl, h=250,forecastfunction=function(x,h,level){
  fit <- Arima(x,order=c(0,0,1),include.mean = F)
  return(forecast(fit,h=250,level=level))})
plot(f,include=1000)
```


Conclusion:
VAR Model does not seem to be a good fit. Need to try VARMA.


